---
title: "R Notebook"
output: html_notebook
---

```{r warning=FALSE, message=FALSE}
#rm(list = ls(all.names = TRUE))
set.seed(132628335)
suppressPackageStartupMessages({
pacman::p_load(tidyverse, reshape2, ggplot2, coda, gtools, gridExtra,
                foreach, doParallel, doRNG, tidyr,plyr, pomp, doMC, zoo, 
                devtools, readxl, viridis, tibbletime, tidyquant, lubridate,
                grid, ggthemes, ggh4x, ggsci, ggnewscale, repr, magrittr, 
               coda, xtable, gghighcontrast, superdiag,ggpubr,cowplot)})

base_colors <- scales::hue_pal()(18)
dark_colors <- colorspace::darken(base_colors, 0.3)
lite_colors <- colorspace::lighten(base_colors, 0.3)

library(doParallel)
registerDoParallel(cores = detectCores())
options(scipen = 100, digits = 4)
```

```{r warning=FALSE, message=FALSE}
set.seed(132628335)
library(tidyverse)
library(here)
library(pomp)
library(foreach)
library(doFuture)
library(parallel)
library(doParallel)
library(doRNG)
library(coda)
library(ggthemes)
library(coda)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggthemes)
library(cowplot)
library(minpack.lm)
library(coda)
library(xtable)
library(mgcv)
library(zoo)

colors = c("#000000","#31E1F7","#F99417","#E11299","#002B5B","#379237","#D2001A","#6F38C5","#8D7B68","#BFDB38","#820000","#367E18","#2192FF","#C58940")
options(scipen = 100, digits = 8)

#####################
# SEROPREVALENCE DATA 
#####################

weekly_seroprevalence <- read.csv("~/Documents/MUN/SEARCH-ID-MODEL/searchid/pomp/data/estimated_seroprevalence.csv")
weekly_seroprevalence$week_end <- as.Date(weekly_seroprevalence$week_end, format = "%Y-%m-%d")

# Make copy data and format the date
original_est_SERP <- weekly_seroprevalence[(weekly_seroprevalence$geo == "NL"),]
original_est_SERP <- original_est_SERP %>%
  mutate(week_end = as.Date(week_end, format = "%Y-%m-%d"),est_infection_date = week_end)

# Subset the Anti-N 
weekly_seroprevalence <- weekly_seroprevalence[(weekly_seroprevalence$geo == "NL")  & (weekly_seroprevalence$ab_estimate == "Anti-N estimate"),] #

# Rename date column
weekly_seroprevalence <- weekly_seroprevalence %>%
  rename_at("week_end", ~"date")

# Format date column
weekly_seroprevalence$date <- as.Date(weekly_seroprevalence$date, format = "%Y-%m-%d")

# Total population
total_population <- 510550

# Create a copy of data
est_SERP <- weekly_seroprevalence

# Create the daily seroprevalence 
est_SERP <- est_SERP %>%
  mutate(daily_pct = c(0,diff(pct_mean)))

# Subset only the study period
est_SERP <- est_SERP[(est_SERP$date >= "2021-12-09") & (est_SERP$date < "2022-09-09"),] #"2022-06-09"
rownames(est_SERP) <- NULL

# est_SERP <- est_SERP %>%
#   select(-c(geo,ab_estimate,pct_q25,pct_q75))

# Create the count index
est_SERP$day <- 1:nrow(est_SERP) * 7
rownames(est_SERP) <- NULL

# create a new dummy dataframe in the daily scale
cum_daily_seroprev <- data.frame(day = 1:(nrow(est_SERP) * 7)) 

# Create a sequence of dates
dates <- seq.Date(from = as.Date("2021-12-11"), by = "day", length.out = nrow(cum_daily_seroprev))

#-----------------------------------------------------------------------------------
# Fit a logistic model to convert weekly estimated seroprevalence to daily estimates
#-----------------------------------------------------------------------------------

# Define the logistic function
# logistic_model <- function(t, K, r, t0) {
#   K / (1 + exp(-r * (t - t0)))
#   }
# # median(est_SERP$day)
# # Fit the serology data on the weekly scale
# logit_model <- nlsLM(pct_mean ~ logistic_model(day, K, r, t0), data = est_SERP,
#              start = list(K = max(est_SERP$pct_mean), r = 0.05, t0 = est_SERP$day[1]),
#              control = nls.lm.control(maxiter = 2000000))
# 
# # Predict cumulative seroprevalence on daily scale
# cum_daily_seroprev$cum_daily_serop <- predict(logit_model, type = "response", newdata = cum_daily_seroprev)

#-----------------------------------------------------------------------------------
# Fit a Gompertz model to convert weekly estimated seroprevalence to daily estimates
#-----------------------------------------------------------------------------------

gompertz_func = function(a,b,c,K, t) {
  a + (c * exp(-exp(-b * (t - K))))
  }
#
gombertz_mod <- nlsLM(pct_mean ~ gompertz_func(a,b,c, K, day), data = est_SERP,
                      start = list(a = 0.5, b = 0.05, c = 1, K = max(est_SERP$pct_mean)),
                      control = nls.lm.control(maxiter = 20000))

y_pred_gomb <- predict(gombertz_mod, newdata = cum_daily_seroprev, type = "response")

cum_daily_seroprev$cum_daily_serop <- y_pred_gomb
#-----------------------------------------------------------------------------------

# Add dates
cum_daily_seroprev$date <- dates

cum_daily_seroprev <- cum_daily_seroprev[(cum_daily_seroprev$date >= "2021-12-15"), ] #& (cum_daily_seroprev$date < "2022-06-03"),
cum_daily_seroprev$day <- 1:nrow(cum_daily_seroprev)
#rownames(cum_daily_seroprev) <- NULL

# After exposure to the COVID-19, it generally takes about two to three weeks for a person to recover and develop antibodies that can be measured in 
# blood tests (Long et al., 2022). We will adjust the seroprevalence data (daily_serop) by shifting the data backward by the average recovery and antibody development period (i.e., 14 days). Implying that cases reported on 2021-12-15 will be detected through antibodies around 2021-12-29. Hence, the seroprevalence will be subsetted from this date and reassigned the date to correspond to the reported cases' date.  

# the lag period in days
lag_days <- 14

# calculate seroprevalence by shifting the daily seroprevalence backward by lag_days
cum_daily_seroprev <- cum_daily_seroprev %>%
  mutate(seroprevalence = lag(cum_daily_serop, lag_days))

# Replace NA values (resulting from the lag) with 0 for visualization purposes
cum_daily_seroprev$seroprevalence[is.na(cum_daily_seroprev$seroprevalence)] <- 0

# subset seroprevalence after shifting forward by 14- days and reassign date to reflect the forward shift
seroprevalence <- cum_daily_seroprev[(cum_daily_seroprev$date >= "2021-12-29"),] #& (cum_daily_seroprev$date < "2022-06-03"),] #"2022-06-09"

#rownames(seroprevalence) <- NULL
# reassign date to reflect the forward shift
date <- seq.Date(from = as.Date("2021-12-15"), by = "day", length.out = nrow(seroprevalence))
seroprevalence$date <- date
seroprevalence <- seroprevalence[(seroprevalence$date < "2022-06-03"),] #"2022-06-09"

seroprevalence$day <- 1:nrow(seroprevalence)
seroprevalence <- subset(seroprevalence, select = c("day","date","seroprevalence"))
rownames(seroprevalence) <- NULL

# ensure date is ordered and derive the daily seroprevalence increment
seroprevalence <- seroprevalence %>%
  arrange(date) %>%
  mutate(daily_serop_increment = c(NA, diff(seroprevalence)))

seroprevalence$daily_serop_increment <- na.approx(seroprevalence$daily_serop_increment)
seroprevalence
```
#-----------------------------------------------------------------------------------
# Add confidence interval (CI) to daily seroprevalence estimates
#-----------------------------------------------------------------------------------
n <- nrow(seroprevalence)
sero_mean_y <- seroprevalence$seroprevalence
sero_sd_y <- sd(seroprevalence$seroprevalence)  # Sample standard deviation

# Z-score for 95% confidence; qnorm(0.975) gives 1.96
z <- qnorm(0.975)
#
# # Calculate confidence interval
sero_ci_width <- z * (sero_sd_y / sqrt(n))
sero_lower_ci <- sero_mean_y - sero_ci_width
sero_upper_ci <- sero_mean_y + sero_ci_width
#
# Add calculated CI to daily seroprevalence estimates
seroprevalence$cum_daily_serop_q025 <- sero_lower_ci
seroprevalence$cum_daily_serop_q975 <- sero_upper_ci

#################
# REPORTED DATA 
#################
# recovered <- read.csv("~/Documents/nl_ba1_two_peak_pmcmc_model/data/covid19-download.csv")
# recovered$date <- as.Date(recovered$date, format = "%Y-%m-%d")
# recovered <- recovered[(recovered$prname == "Newfoundland and Labrador"),] #
# recovered <- recovered[,c("date","numrecover")]
# rownames(recovered) <- NULL
# recovered <- recovered[(recovered$date >= "2021-12-15") & (recovered$date < "2022-06-03"),]
# recovered$daily_recovery <- c(0,diff(recovered$numrecover))
# recovered$recovery <- cumsum(recovered$daily_recovery) / total_population

#################
# REPORTED DATA 
#################

# load reported data
reported_data <- read.csv("~/Documents/MUN/SEARCH-ID-MODEL/searchid/pomp/data/IM215194_NL_COVID_CASESHOSPDEATH.csv")

# format date column
reported_data$DATE <- as.Date(reported_data$DATE, format = "%Y-%m-%d")

# rename columns
colnames(reported_data) <- c("date", "cases", "hosp", "death")

# reassign row indices
rownames(reported_data) <- 1:nrow(reported_data)

# select only date and cases columns
reported_data <- reported_data[c("date", "cases")]

# Calculate daily cumulative cases
reported_data$cum_cases <- cumsum(reported_data$cases)

# Calculate total cases, last date, weekly cumulative sum, and overall cumulative sum
reported_data$cum_case_per_capita <- reported_data$cum_cases / total_population

# rearrange data according to date in ascending order
reported_data <- reported_data %>%
  arrange(date)

# Calculate total cases, last date, weekly cumulative sum, and overall cumulative sum
reported_data$daily_case_increment = c(NA, diff(reported_data$cum_case_per_capita))

# Calculate total cases, last date, weekly cumulative sum, and overall cumulative sum
# Subset data for study period
reported_data <- reported_data[(reported_data$date >= "2021-12-04") & (reported_data$date < "2022-06-03"),]

# create a new column to count days
reported_data$day <- 1:nrow(reported_data)

# remove row indices
rownames(reported_data) <- NULL
# plot(reported_data$day,reported_data$daily_case_increment)
# plot(cum_daily_seroprev$day,cum_daily_seroprev$daily_serop_increment)

#-----------------------------------------------------------------------------------
# Add confidence interval (CI) to daily case per capita estimates
#-----------------------------------------------------------------------------------

n <- nrow(reported_data)
case_mean_y <- reported_data$cum_case_per_capita
case_sd_y <- sd(reported_data$cum_case_per_capita)  # Sample standard deviation

# Z-score for 95% confidence; qnorm(0.975) gives 1.96
z <- qnorm(0.975)

# Calculate confidence interval
case_ci_width <- z * (case_sd_y / sqrt(n))
case_lower_ci <- case_mean_y - case_ci_width
case_upper_ci <- case_mean_y + case_ci_width
# 
# Add calculated CI to daily seroprevalence estimates
#daily_reported_cases$daily_mean_case_capita <- case_mean_y
reported_data$cum_case_per_capita_q025 <- case_lower_ci
reported_data$cum_case_per_capita_q975 <- case_upper_ci

# subset the data from 2021-12-15 to 2022-06-02 and with only required columns
reported_data <- reported_data[(reported_data$date >= "2021-12-15") & (reported_data$date < "2022-06-03"),] #"2022-06-09"
reported_data <- subset(reported_data, select = c("day", "date", "cases"))
reported_data$day <- 1: nrow(reported_data)
rownames(reported_data) <- NULL
seroprevalence
reported_data
```
# cbind only selected columns from seroprevalence
merged_data <- cbind(reported_data, seroprevalence[c("seroprevalence","daily_serop_increment")])

# now we are ready to create the adjusted cases column using the daily seroprevalence and total population
merged_data$adjusted_serop_cases <- as.integer(merged_data$daily_serop_increment * 510550)

plot(merged_data$adjusted_serop_cases)
```
# Create a full join and merge together using the new weekly dates
# merged_data <- seroprevalence %>%
#   full_join(reported_data, by = c("date","day"))

# merged_data <- merged_data %>%
#   full_join(recovered[,c("date","recovery")], by = "date")

# # Estimate underreporting factor
merged_data$underreporting_factor <- merged_data$daily_serop_increment / merged_data$daily_case_increment
# plot(merged_data$day,merged_data$cum_daily_serop / merged_data$cum_case_per_capita)

#-----------------------------------------------------------------------------------
# Add confidence interval (CI) to underreporting estimate
#-----------------------------------------------------------------------------------

n <- nrow(merged_data)
underrep_mean_y <- merged_data$underreporting_factor
underrep_sd_y <- sd(merged_data$underreporting_factor)  # Sample standard deviation

# Z-score for 95% confidence; qnorm(0.975) gives 1.96
z <- qnorm(0.975)

# Calculate confidence interval
underrep_ci_width <- z * (underrep_sd_y / sqrt(n))
underrep_lower_ci <- underrep_mean_y - underrep_ci_width
underrep_upper_ci <- underrep_mean_y + underrep_ci_width
# 
# Add calculated CI to daily seroprevalence estimates
# merged_data$daily_mean_underrep <- underrep_mean_y
merged_data$daily_underrep_q025 <- underrep_lower_ci
merged_data$daily_underrep_q975 <- underrep_upper_ci

# Convert underreporting rates to actual numbers to estimate the true infections in the population
merged_data$estimated_true_infections <- as.integer(merged_data$cases * merged_data$underreporting_factor)
rownames(merged_data) <- NULL 

# write.csv(merged_data, file = "~/Documents/nl_ba1_two_peak_pmcmc_model/data/omicron_sero_case_estimated_infections.csv", row.names = FALSE)
merged_data

plot(merged_data$daily_serop_increment * 510550)
plot(merged_data$cases)
plot(merged_data$cum_daily_serop)

```